### 2016-10-14 dissertation/normal.tex
dissertation part normal: update 6

\section{Normal Basis Multiplier over Galois Field}
\subsection{Normal Basis}
Given a Galois field (GF) $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

@@

Mastrovito multiplier and Montgomery multiplier are 2 common designs
of GF multipliers using StdB. As a naive implementation of GF multiplication,
Mastrovito multiplier uses most number of gates:
$k^2$ AND gates plus $k^2-\Delta$ XOR gates \cite{Mastrovito}. Montgomery multiplier 
applies lazy reduction techniques and results in a better latency performance, while the number of gates are about
the same with Mastrovito multiplier:
$k^2$ AND gates plus $k^2-k/2$ XOR gates \cite{Montgomery}. 
Concretely, typical design of Mastrovito multiplier consists of 218 logic gates, while 
Montgomery multiplier needs 198 gates. However, the NB multiplier reuses the $\lambda$-Matrix 
logic, so this component will only need to be implemented for once. 
Consider the definition of matrix multiplication, it needs $C_N$ AND gates to apply 
bit-wise multiplication and $C_N-1$ XOR gates to sum them up. The number of nonzero entries
in the $\lambda$-Matrix can be counted: $C_N = 27$.
As a result, the most naive NB multiplier design (or Massey-Omura multiplier \cite{MasseyOmura})
contains 53 gates in total, which is a great saving in area cost comparing to StdB multipliers.

>>>>

\section{Normal Basis Multiplier over Galois Field}
From algebraic view, a field is a space, and field elements are dots in the space. Those elements can be 
represented with unique coordinates, which requires the pre-definition of a basis vector. In this
section, we discuss a special basis called normal basis, as well as the advantages adopting it in 
GF operations esp. multiplication.
\subsection{Normal Basis}
Given a Galois field (GF) $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

@@

Mastrovito multiplier and Montgomery multiplier are 2 common designs
of GF multipliers using StdB. As a naive implementation of GF multiplication,
Mastrovito multiplier uses most number of gates:
$k^2$ AND gates plus $k^2-\Delta$ XOR gates \cite{Mastrovito}. Montgomery multiplier 
applies lazy reduction techniques and results in a better latency performance, while the number of gates are about
the same with Mastrovito multiplier:
$k^2$ AND gates plus $k^2-k/2$ XOR gates \cite{Montgomery}. 
Concretely, typical design of Mastrovito multiplier consists of 218 logic gates, while 
Montgomery multiplier needs 198 gates. However, the NB multiplier reuses the $\lambda$-Matrix 
logic, so this component will only need to be implemented for once. 
Consider the definition of matrix multiplication, it needs $C_N$ AND gates to apply 
bit-wise multiplication and $C_N-1$ XOR gates to sum the intermediate products up. The number of nonzero entries
in the $\lambda$-Matrix can be counted: $C_N = 27$.
As a result, the most naive NB multiplier design (or Massey-Omura multiplier \cite{MasseyOmura})
contains 53 gates in total, which is a great saving in area cost comparing to StdB multipliers.

\section{Design a Normal Basis Multiplier on Gate Level}
The NB multiplier design consumes much less gates than ordinary StdB multiplier design, even if 
we use the most naive design. However, the modern NB multiplier design has been improved a lot from the 
very first design model proposed by Massey and Omura in 1986 \cite{MasseyOmura}. In order to 
test our approach on practical contemporary circuits, it is necessary to learn the mechanism and design 
routine of several kinds of modern NB multipliers.
\subsection{Sequential Multiplier with Parallel Outputs}
The major benefit of NB multiplier origins from the sequential design. A straightforward design implementing 
the cyclic-shift of operands and $\lambda$-Matrix logic component is the Massey-Omura multiplier.

% fig:MasseyOmura

Figure \ref{fig:MasseyOmura} shows the basic architecture of a Massey-Omura multiplier. The operands
$A$ and $B$ are 2 arrays of flip-flops which allow 1-bit right-cyclic shift every clock cycle.
The logic gates in the boxes implements the matrix multiplication with $\lambda$-Matrix $M^{(0)}$, 
while each AND gate corresponds to term $a_ib_j$ and each XOR gate corresponds to addition $a_ib_j+a_{i'}b_{j'}$. 
The XOR layer has only 1 output, giving out 1 bit of product $C$ every clock cycle.

The behavior of Massey-Omura multiplier can be concluded as: pre-load operands $A,B$ and reset $C$ to 0, after 
executing for $k$ clock cycles, the data stored in flip-flop array $C$ is the product $A\times B$.
We note that there is only one output giving 1 bit of the product each clock cycle, which matches the 
definition of serial output to communication channel. Therefore this type of design is named as 
sequential multiplier with serial output (SMSO).
The SMSO architecture need $C_N$ AND gates and $C_N-1$ XOR gates, which equals to $2k-1$ AND gates and $2k-2$ XOR gates
if it is designed using ONB. In fact, the number of gates can be reduced if the multiplication is 
implemented using a conjugate of SMSO.

The gate-level logic boxes are implementing following function:
\begin{equation}
\label{eqn:SMPOterms}
c_l = row_1(A\times M^{(l)}) \times B + row_2(A\times M^{(l)}) \times B + \cdots + row_{k}(A\times M^{(l)}) \times B
\end{equation}
It can be decomposed into $k$ terms. If we only compute one term for each $c_l,~0\leq l\leq k-1$ in one 
clock cycle, make $k$ outputs and add them up using shift register after $k$ clock cycles, 
it will generate the same result with SMSO. This kind of architecture is named as 
sequential multiplier with parallel outputs (SMPO). The basic SMPO, as a conjugate of Massey-Omura multiplier,
is invented by Agnew et al. \cite{agnew1991implementation}.

\begin{figure}[H]
\centering{
\includegraphics[width=\textwidth]{newfig/mySMPO.eps}
\caption{5-bit Agnew's SMPO. Index $i$ satisfies $0<i<4$, indices $u,v$ are determined by column \# of nonzero entries in $i$-th row of $\lambda$-Matrix $M^{(0)}$, i.e. if entry $M_{ij}^{(0)}$ is a nonzero entry, $u$ or $v$ equals to $i+j \pmod 5$. Index $w = 2i\pmod 5$}
\label{fig:SMPO}}
\end{figure}

\begin{Example}[5-bit Agnew's SMPO]
Given GF $\mathbb F_{2^5}$ and primitive element $\alpha$ defined by irreducible polynomial 
$\alpha^5+\alpha^2+1=0$, normal element $\beta = \alpha^5$ constructs an ONB $\{\beta,\beta^2,\beta^4,\beta^8,\beta^{16}\}$.
The $0$-th $\lambda$-Matrix for this ONB is
\begin{equation*}
M^{(0)} = \left(\begin{array}{lcccr}
0 &1 &0 &0 &0 \\
1 &0 &0 &1 &0 \\
0 &0 &0 &1 &1 \\
0 &1 &1 &0 &0 \\
0 &0 &1 &0 &1
\end{array}\right)
\end{equation*}
Then a typical design of 5-bit Agnew's SMPO is depicted in figure \ref{fig:SMPO}.

The operands part of this circuit is the same with Massey-Omura multiplier. The differences are on 
the matrix multiplication part, while it is implemented as separate logic blocks for 5 outputs,
and the 5 blocks are connected in a shift register fashion. By analyzing the detailed function of 
logic blocks, we can reveal the mechanism of Agnew's SMPO.

Suppose we implement $M^{(0)}$ as the logic block in SMSO. In the first clock cycle, the output is
\begin{equation}
\label{eqn:5bitSMPO_c0}
c_0 = a_1b_0+(a_0+a_3)b_1 + (a_3+a_4)b_2 + (a_1+a_2)b_3 + (a_2+a_4)b_4
\end{equation}
Note it is written in the form of equation \ref{eqn:SMPOterms}. In next clock cycles we can obtain 
remaining bits of the product, which can be written in following general form polynomial:
\begin{align}
c_i =& b_ia_{i+1} + b_{i+1}(a_i + a_{i+3}) + b_{i+2}(a_{i+3} + a_{i+4}) \nonumber\\
&+ b_{i+3}(a_{i+1} + a_{i+2}) + b_{i+4}
(a_{i+2} + a_{i+4}),\ 0\leq i\leq 4 \nonumber
\end{align}
Note all index calculations are reduced modulo 5.

Now let us observe the behavior of 5-bit Agnew's SMPO. Initially all DFFs are reset to 0. 
In the first clock cycle,
signal sent to the flip-flop in block $R_0$ denotes function:
$$R_0^{(1)} = a_1b_0$$
It equals to the first term of equation \ref{eqn:5bitSMPO_c0}. In the second clock cycle,
this signal is sent to block $R_1$ through wire $r_0$, and this block also receives data 
from operands (shifted by 1 bit), generating signal $a_u,a_v$ and $b_w$. Concretely,
signal sent to flip-flop in block $R_1$ is:
$$R_1^{(2)} = R_0^{(1)} + (a_0+a_3)b_1 = a_1b_0+(a_0+a_3)b_1$$
which forms first 2 terms of equation \ref{eqn:5bitSMPO_c0}. Similarly, we track the signal 
on $R_2$ in third clock cycle, signal on $R_3$ in fourth clock cycle, finally we can get 
$$R_4^{(5)} = a_1b_0+(a_0+a_3)b_1 + (a_3+a_4)b_2 + (a_1+a_2)b_3 + (a_2+a_4)b_4$$
which equals to $c_0$ in equation \ref{eqn:5bitSMPO_c0}.
After the fifth clock cycle ends, this signal can be detected on wire $r_0$. It shows that 
the result of $c_0$ is computed after 5 clock cycles and gives on $r_0$.

If we track $R_1\to R_2\to R_3 \to \cdots \to R_0$, we can obtain $c_1$ respectively.
Thus we conclude that Agnew's SMPO functions the same with Massey-Omura multiplier.
\end{Example}

The design of Agnew's SMPO guarantees that there is only one AND gate in each $R_i$ block.
For ONB, adopting Agnew's SMPO will reduce the number of AND gates from $2k-1$ to $k$.

\subsection{Multiplier not based on $\lambda$-Matrix}
